The Choose-Your-Own-Adventure Calculus (Pearl/Brave New Idea)

### 2025-02-16 paper.tex
The Gamma example

\newcommand{\ident}[1]{\textsf{#1}}
\newcommand{\sigmaN}{\sigma_0}
\newcommand{\select}{\textnormal{\ident{choose}}}
\newcommand{\preview}{\textnormal{\ident{preview}}}
\newcommand{\links}{\textnormal{\ident{links}}}
\newcommand{\choices}{\textnormal{\ident{choices}}}
\newcommand{\op}{\textit{op}}
\newcommand{\argmax}{\textit{arg}\,\textit{max}\,}

@@

\subparagraph{Type providers}

@@

\subparagraph{Data exploration}

@@

\subparagraph{AI assistants}

@@

\subparagraph{Interactive theorem provers}
A fourth example of the choose-your-own-adventure interactive pattern can be found in interactive
theorem provers. When writing programs in systems like Idris \cite{brady-2021-idris2}, the
user typically works by stating the desired conclusion and filling the implementation with a hole.
The system provides a range of interactive editing capabilities to fill the holes \cite{mcbride-1999-dependently}.
It can, for example, generate a case split or search for a proof \cite{brady-2015-idris}.

@@

\vspace{-0.25em}
\begin{itemize}
  \item $\choices(\sigma) = (\iota_1\mapsto\sigma_1, \ldots, \iota_n\mapsto\sigma_n)$ is
    an operation that takes a state and generates options designated by an identifier $\iota_i$
    and represented by a state $\sigma_i$,
  \item $\select(\sigma) &=& e$ is an operation that returns generated program for a given state.
\end{itemize}
\end{definition}

@@

\subparagraph{External mode of embedding}
One of the subtle questions about the choose-your-own-adventure pattern raised in the introduction
concerns the different ways in which a trace of the interaction is embedded in the interactively
constructed program. In the \emph{external mode}, the interaction results in code that becomes a
part of the edited program, but it is not possible to reconstruct the steps used to generate the code.

@@

\subparagraph{Internal mode of embedding}
In the \emph{internal mode}, a trace of the interaction with a choose-your-own-adventure system
is embedded directly in the constructed program. This is the case with type providers, where a
user chooses a sequence of object members to be accessed. The same would be the case in a completion
system for Coq that would offer tactics to apply, becuase the resulting proof would contain a
record of the selected tactics.

@@

\section{Examples}
\label{sec:examples}

@@

@@

\subsection{Interactive theorem proving}

>>>>

\newcommand{\ident}[1]{\textsf{#1}}
\newcommand{\ddident}[1]{\guillemotleft\ident{#1}\guillemotright}
\newcommand{\sigmaN}{\sigma_0}
\newcommand{\select}{\textnormal{\ident{choose}}}
\newcommand{\choices}{\textnormal{\ident{choices}}}
\newcommand{\op}{\textit{op}}
\newcommand{\vect}[1]{\lbrack #1 \rbrack}
%\newcommand{\preview}{\textnormal{\ident{preview}}}
%\newcommand{\links}{\textnormal{\ident{links}}}
%\newcommand{\argmax}{\textit{arg}\,\textit{max}\,}

@@

\subparagraph{Type providers.}

@@

\subparagraph{Data exploration.}

@@

\subparagraph{AI assistants.}

@@

\subparagraph{Interactive theorem provers.}
A fourth example of the choose-your-own-adventure interactive pattern can be found in interactive
theorem provers. When writing programs in systems like Idris \cite{brady-2021-idris2}, the
user typically works by stating the desired conclusion and filling the implementation with a hole.
The system provides a range of interactive editing capabilities to fill the holes \cite{mcbride-1999-dependently}.
It can, for example, generate a case split or search for a proof \cite{brady-2015-idris}.

@@

\vspace{-0.25em}
\begin{itemize}
  \item $\choices(\sigma) = (\iota_1\mapsto\sigma_1, \ldots, \iota_n\mapsto\sigma_n)$ is
    an operation that takes a state and generates options designated by an identifier $\iota_i$
    and represented by a state $\sigma_i$,
  \item $\select(\sigma) = e$ is an operation that returns generated program for a given state.
\end{itemize}
\end{definition}

@@

\subparagraph{External mode of embedding.}
One of the subtle questions about the choose-your-own-adventure pattern raised in the introduction
concerns the different ways in which a trace of the interaction is embedded in the interactively
constructed program. In the \emph{external mode}, the interaction results in code that becomes a
part of the edited program, but it is not possible to reconstruct the steps used to generate the code.

@@

\subparagraph{Internal mode of embedding.}
In the \emph{internal mode}, a trace of the interaction with a choose-your-own-adventure system
is embedded directly in the constructed program. This is the case with type providers, where a
user chooses a sequence of object members to be accessed. The same would be the case in a completion
system for Coq that would offer tactics to apply, becuase the resulting proof would contain a
record of the selected tactics.

@@

\newpage
\section{Examples}
\label{sec:examples}

We now revisit the four examples from Section~\ref{sec:motivation} and show how they fit the
above formal model. All four examples rely on some domain-specific logic. We describe what
information the logic provides, but do not model it formally. This has been done elsewhere,
in works describing the individual systems.

To show how the model lets us distinguish subtle details of interactive programming systems,
we start with a model of data exploration system that is inspired by The Gamma, but differs in
one notable way. We then discuss type providers more generally and show how to correctly model
The Gamma. We then revisit the remaining two examples.

In The Gamma, the choose-your-own-adventure interaction pattern is used to construct a
query that transforms the given input data. The query is a sequence of operations with parameters,
$\op(p_1, \ldots, p_n)$, loosely modelled after relational algebra \cite{codd-1970-relational}.

In The Gamma, the query is hidden from the data analyst. Behind the scenes, the system generates
objects with members and the identifiers designating individual options are the names of those
members. The operation is encapsulated in the code of the accessor of the member. In the
simplified model in this section we ignore this fact. The model presented here directly
generates code that calls the underlying operations. For example, assume that the user makes
the following choices:
\[
\textnormal{\ddident{group data}\,.\,\ddident{by Athlete}\,.\,\ddident{sum Gold}\,.\,\ddident{count all}\,.\,\ddident{then}}
\]
In The Gamma, the individual identifiers become object members and they are included as
a member chain in the generated code. In the following simplified model, the completion
instead fills the hole with an expression representing the operation:
\[
\ident{group}(\texttt{"Athlete"}, \ident{sum}(\texttt{"Gold"}), \ident{count}())
\]
The two approaches have different human-computer interaction trade-offs. In terms of cognitive
dimensions \cite{green-1989-cogdims,blackwell-2003-cogdims}, the latter has a greater closeness
of mapping, while the former is less cognitively demanding to read for a non-programmer.
As discussed in Section~\ref{sec:properties}, the two implementations of the choose-your-own-adventure
interaction pattern also differ in terms of their formal properties.

\subparagraph{Formal model.}
The options generated by The Gamma let the user select both the next operation and the parameters
of the previously selected operation. The available operations and parameters are generated
based on a schema $S$ that is transformed by the operations. The state of the system $\sigma$
contains the current schema $S$ and the operations applied so far. In the following, we write
$\op(\boldsymbol{p})$ for an operation with a vector of parameters:
\[
\sigma = S, \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n})}\\
\]
The behaviour of the $\choices$ operation depends on whether the last operation in the sequence
expects further parameters or whether it is fully-specified. In the first case, the recommendation
engine generates possible additional parameter values $p_', p_'', \ldots$ based on the schema $S$,
the operation $\op_n$ and the already known parameters $\boldsymbol{p_n}$.
The $\choices$ operation then generates options that add the additional parameter. We generated the
identifiers $\iota_',\iota_'',\ldots$ based on the state and the parameter value, such as
\ddident{by Gold descending}. Note that adding a parameter may also result in a new schema
$S', S'', \ldots$ (which the recommendation engine computes based on the previous schema and the new parameter):
\[
\begin{array}{l}
\choices(S, \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n})}) =\\
\qquad (S', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n},p')}), \\
\qquad (S'', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n},p'')}), ~\ldots
\end{array}
\]
If the last operation takes no further parameters, the system produces a choice
of possible next operations $\op', \op'', \ldots$. Again, we are also given new schemas $S', S'', \ldots$
and we generate identifiers $\iota',\iota'',\ldots$ based on the operation name. The $\ident{choices}$
operation then returns options that add the additional operation:
\[
\begin{array}{l}
\choices(S, (\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}))) =\\
\qquad (S', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}), \op'()}), \\
\qquad (S'', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}), \op''()}), ~\ldots
\end{array}
\]
Finally, the $\select$ operation takes the state $\sigma$ and generates an expression that represents
the data transformation. This is only possible if all parameters are fully-specified. For simplicity,
assume that $k$ is the index of the last fully-specified operation (either $n$ or $n-1$). If the
host language lets us compose functions using $f\circ g$, we can write:
\[
\select(S, (\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}))) = \op_1(\boldsymbol{p_1}) \circ \ldots \circ \op_k(\boldsymbol{p_k})
\]
The recommendation engine behind The Gamma provides a domain-specific logic for generating
possible operations and their parameters based on the current schema of the data. As the above
definition shows, this underlying engine can be easily exposed through the common
choose-your-own-adventure interface.

\subsection{Type providers}

revisit

@@

\subsection{Interactive theorem proving}
~

@@
